---
title: "Beverton-Holt aquatic model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beverton-Holt aquatic model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicroMoB)
library(ggplot2)
library(data.table)
library(parallel)
```

The simplest non-trivial model of aquatic mosquito dynamics which exhibits density-dependence (negative feedback)
is a Beverton-Holt (BH) style model. It is a kind of discrete logistic equation. We have implemented a BH model that
conforms to the aquatic (immature) component interface to be used with other mosquito models fulfilling the adult
component interface.

The model groups all immature mosquito stages into a single compartment `L`. Each time step some number
may mature and be ready to emerge as adults, `A`. The model requires 3 (potentially time and patch varying parameters),
`K`, which determines the strength of density dependent mortality, `surv`, the baseline daily survival probability,
and `molt`, the daily proportion of immature mosquitoes which mature and emerge as adults.

## Deterministic model

The state updating equation is as follows:

\begin{equation}
L_{t+1} = eggs + (1-molt) \cdot surv \cdot L_{t} \left( \frac{K}{L_{t}+K} \right) \\
A_{t} = molt \cdot surv \cdot L_{t} \left( \frac{K}{L_{t}+K} \right)
\end{equation}

## Stochastic model

In the stochastic model on a time step, we first draw the number of survivors:

\begin{equation}
survivors \sim \mathrm{binomial}(L, surv \left( \frac{K}{L_{t}+K} \right)
\end{equation}

The number who emerge is also a binomial random variate:

\begin{equation}
emerge \sim \mathrm{binomial}(survivors, molt)
\end{equation}

Then state is updated:

\begin{equation}
L_{t+1} = eggs + survivors - emerge \\
A_{t} = emerge
\end{equation}

## Deterministic equilbrium

Despite being a nonlinear equation, we can solve the equilibrium. We assume that `eggs` is known, and that the desired
daily emergence is known (`lambda`, or `A` in the BH model's notation). Then we solve for `K` and `L`:

\begin{equation}
L = \lambda \left( \frac{1}{molt} - 1 \right) + eggs \\
K = \frac{-\lambda L}{\lambda - L \cdot molt \cdot surv}
\end{equation}

We can set up a simulation using the RM mosquito model to check that everything is
at equilibrium. First we calculate the equilibrium values. We do not use any components
other than the aquatic and adult mosquito components.

```{R}
patches <- 1
tmax <- 1e2

M <- 120
p <- 0.9
lambda <- M*(1-p)

nu <- 25
f <- 0.3
eggs <- nu * f * M

# static pars
molt <-  0.1
surv <- 0.9

# solve L
L <- lambda * ((1/molt) - 1) + eggs
K <- - (lambda * L) / (lambda - L*molt*surv)
```

Now we run a deterministic simulation and draw 10 stochastic trajectories.

```{R}
# deterministic run
mod <- make_MicroMoB(tmax = tmax, p = patches)
setup_aqua_BH(model = mod, stochastic = FALSE, molt = molt, surv = surv, K = K, L = L)
setup_mosquito_RM(model = mod, stochastic = FALSE, f = f, q = 0.9, eip = 10, p = p, psi = diag(1), nu = nu, M = M, Y = 0, Z = 0)

out_det <- matrix(data = NaN, nrow = tmax + 1, ncol = 3)
out_det[1L, ] <- c(mod$aqua$L, mod$aqua$A, mod$mosquito$M)

while (mod$global$tnow <= tmax) {
  step_aqua(model = mod)
  step_mosquitoes(model = mod)
  out_det[mod$global$tnow + 1L, ] <- c(mod$aqua$L, mod$aqua$A, mod$mosquito$M)
  mod$global$tnow <- mod$global$tnow + 1L
}

# stochastic runs
out_sto <- mclapply(X = 1:10, FUN = function(runid) {
  
  mod <- make_MicroMoB(tmax = tmax, p = patches)
  setup_aqua_BH(model = mod, stochastic = TRUE, molt = molt, surv = surv, K = K, L = L)
  setup_mosquito_RM(model = mod, stochastic = TRUE, f = f, q = 0.9, eip = 10, p = p, psi = diag(1), nu = nu, M = M, Y = 0, Z = 0)
  
  out_run <- matrix(data = NaN, nrow = tmax + 1, ncol = 3)
  out_run[1L, ] <- c(mod$aqua$L, mod$aqua$A, mod$mosquito$M)
  
  while (mod$global$tnow <= tmax) {
    step_aqua(model = mod)
    step_mosquitoes(model = mod)
    out_run[mod$global$tnow + 1L, ] <- c(mod$aqua$L, mod$aqua$A, mod$mosquito$M)
    mod$global$tnow <- mod$global$tnow + 1L
  }
  
  out_run <- as.data.frame(out_run)
  out_run$run <- as.integer(runid)
  return(out_run)
})


out_det <- as.data.table(out_det)
out_det[, "Day" := 0:tmax]
out_det <- melt(out_det, id.vars = "Day", variable.name = "Stage", value.name = "Count")
levels(out_det$Stage) <- c("L", "A", "M")

out_sto <- do.call(rbind, out_sto)
out_sto <- as.data.table(out_sto)
out_sto <- melt(out_sto, id.vars = "run", variable.name = "Stage", value.name = "Count")
out_sto[, "Day" := 0:tmax, by = c("run", "Stage")]
levels(out_sto$Stage) <- c("L", "A", "M")

ggplot(data = out_sto) +
    geom_line(aes(x = Day, y = Count, color = Stage, group = run), alpha = 0.35) +
    geom_line(data = out_det, mapping = aes(x = Day, y = Count, color = Stage)) +
    facet_wrap(. ~ Stage, scales = "free")
```
