---
title: "key-themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{key-themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicroMoB)
```

Here we are going to talk about some key themes to follow while building this software library.

## Data provenance

It should be clear how external input data gets turned into the parameters used
by Micro-MoB during simulation. Specifically, if we define some tree 
structure of data such that the leaves are mutually exclusive (e.g.; entomology,
vector control, human movement, drug treatment, etc), we know for each
data structure and function what leaves they use.

This is important because it lets us figure out what changes when data changes.

## Model structure

We need to get a list of model state objects, and updating functions. Then we need
to describe how the process executes a time step. This helps us uncover hidden
constraints in the model.

This is also a step closer to being able to construct multiple representations of
the model (e.g. as machines or resource sharers).

## Extensions

It needs to be clear how to link to this package from another package if you want
to extend the methods here. We need to think about what kind of a simulation
engine we are making and how other people will extend it.

One thing we can do here is make the basic state objects (see above topic) and 
define how to queue updates and access them, and a simulation loop that updates
them, then users can write their own processes. Like what we do with individual
but with more enforced constraints.

## Software design

Once we establish what are the data structures that store persistent state in the
model, we need to decide how to enforce some kind of type consistency for them.
This assumes that those data structures will be used by *all* models that can
be built using Micro-MoB.

One way is to use assertions prior to calling `UseMethod` to make sure that the 
required inputs are provided.

```r
BloodFeeding = function(t, Mod){
  UseMethod("BloodFeeding", Mod$bloodfeedingPars)
}
```

This is probably the right way to do things. The outer generic function can just
check that some bare minimum of input exists. The implemented specific function
can do further checking based on its internal state representation.

Another way would be more complex. We could enforce that users interact with the model
object through an interface which ensures consistency.

## Model object

We should store the model object as an environment. [This gist](https://gist.github.com/slwu89/a294203e6d6679b923276030516684d7)
shows why. Luckily conversion to-from is as easy as `list2env` and `as.list`.

# Simulation

## Simulation loop

Here it is:

```r
MicroSimulation = function(Model){with(Model$runtime,{
  # Defined in Micro-MoB
  for(t in T0:Tmax){
    if (t %% 100 == 0) print(paste("t =", t)) 
    Model = dMicroSim(t, Model)
    Model = IO(t, Model)
    Model = Update(t, Model)
  } 
  return(Model)
})}
```

## Update

There's a couple of ways to do a discrete time simulation, regardless of if
it is seen as a discretization of an underlying continuous time process or the
model is developed "from scratch" as a discrete time model (this amounts to
asking if a rate or a probability is the fundamental quantity specified by a modeler).

Anyway for discrete time, we basically can either do:

  1. Synchronous update - this is like how cellular automata work (usually), all
  units which may undergo a state change draw their state changes simultaneously
  based on the same (yet unchanged) state. All updates are then applied instantaneously.
  This is what individual does, even though it doesn't initially appear to be the
  case, because updated are merely queued by processes, they aren't applied until
  the time step ends.
  2. Asynchronous update - this is what Micro-MoB does. State is updated by processes
  which are executed during each time step. This requires more caution, although
  it can result in processes functions which are shorter and more specialized.
  Basically, state that is used to compute probabilities or proportions _cannot_
  be modified until it is no longer being used to compute those dynamic values.

The former means you need to spend more time working on the model's engine ... its
software architecture. That's why individual has such a large amount of code dedicated
to queuing and executing state updates. So you spend some upfront cost, but you
get to have more leeway with your model processes.

The latter is easier to get running with right away and is a bit more freewheeling.
But you need to be very careful about the execution order of processes. The upshot
is that it may be easier to compartmentalize functions that work on smaller parts
of state.

The terms synchronous and asynchronous are wrong, because these mean quite different
things [when used rigorously](https://en.wikipedia.org/wiki/Asynchronous_cellular_automaton).
We're just talking about queuing updates that execute all at once versus
executing updates sequentially in such a way that it mimics queuing those updates.

The update loop is fundamental. Let's see what it says:

```r
dMicroSim = function(t, Mod){
  # Defined in Micro-MoB
  ##########################################
  # + IRS Spraying
  # + ITN Mass Distribution
  # ... and other modes of vector control
  # + Mass Treatment
  # + Reactive Case Detection
  # ... and other health-based interventions
  ##########################################
  Mod = Intervene(t, Mod)
  
  ##########################################
  #  Called by: 
  #    + some models of aquatic mosquito
  #      populations with dynamic resources 
  #      & resource-based competition. 
  #########################################
  #Mod = Update(t, Mod)  

  ###########################
  # Micro-Human.Rmd 
  # + TimeSpent
  # + MyITN, UseITN
  ###########################
  Mod = HumanBehavior(t, Mod)

  ###########################
  # VectorControl.Rmd 
  # + ITNs, IRS, LSM
  ###########################
  Mod = VectorControlEffects(t, Mod)

  #######################
  # Micro-BloodFeeding.Rmd
  # + FeedingSuccess modifies MPars$psi
  # + kappa, eir, Q
  #######################
  Mod = BloodFeeding(t, Mod)
  
  ##############################
  # Micro-Mosquito.Rmd
  # Micro-Mosquito-Adult.Rmd
  # Micro-Mosquito-Aquatic.Rmd
  # + dMdt
  # + dLdt
  ##############################
  Mod = MosquitoDynamics(t, Mod) 

  ###############################
  # HumanMicro.Rmd
  # + P.falciparum
  ###############################
  Mod = InfectionDynamics(t, Mod)
  
  return(Mod) 
}
```

So here's the order in which things happen:

  1. `Intervene` - this instantaneously changes model state before anything happens. In this
  sense it happens "between" the ticks of the clock, so that all updates see the newly intervened-upon
  state when they execute their updates on this day.
  2. 
