---
title: "key-themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{key-themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MicroMoB)
```

Here we are going to talk about some key themes to follow while building this software library.

## Data provenance

It should be clear how external input data gets turned into the parameters used
by Micro-MoB during simulation. Specifically, if we define some tree 
structure of data such that the leaves are mutually exclusive (e.g.; entomology,
vector control, human movement, drug treatment, etc), we know for each
data structure and function what leaves they use.

This is important because it lets us figure out what changes when data changes.

## Model structure

We need to get a list of model state objects, and updating functions. Then we need
to describe how the process executes a time step. This helps us uncover hidden
constraints in the model.

This is also a step closer to being able to construct multiple representations of
the model (e.g. as machines or resource sharers).

## Extensions

It needs to be clear how to link to this package from another package if you want
to extend the methods here. We need to think about what kind of a simulation
engine we are making and how other people will extend it.

One thing we can do here is make the basic state objects (see above topic) and 
define how to queue updates and access them, and a simulation loop that updates
them, then users can write their own processes. Like what we do with individual
but with more enforced constraints.
